\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-26 paper/algorithm.tex
paper: \ref to \Cref for standardized section names

Our optimizer searches (Sec.~\ref{sec:planning}) a restricted space (Sec.~\ref{sec:search-space}) of query plans containing imputation operators (Sec.~\ref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.

This plan is subject to two constraints (Sec.~\ref{sec:placement}).
First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
Our planning algorithm is agnostic to the type of imputation used (Sec.~\ref{sec:imputation}), as the properties of the imputation are summarized by the cost model (Sec.~\ref{sec:cost-model}).

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in Sec.~\ref{sec:imputation}).
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

@@

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
Figure~\ref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.

@@

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans. 
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the base implementation of our system,
we use an iterative algorithm (see Sec.~\ref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

@@

The query planner must select a join ordering in addition to placing imputation operators as described in Section~\ref{sec:placement}.

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes and a choice over the plan frontier
can only be made once the set is final. Details are shown in Figure~\ref{fig:semantics-plan-cache}.

@@

Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top-level planner, shown in Algorithm~\ref{algo:top-level-planner}.

The planner operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (Lines~\ref{lst:line:attr-start}-\ref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache (Line~\ref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (Line~\ref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (Line~\ref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (Line~\ref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

@@

Theorem~\ref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in Figure~\ref{fig:bad-sit-diagram}

@@

% arrows
\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
\end{tikzpicture}
\caption{A bad Pareto frontier selection would choose $p_2$ and never reach $p_1'$}
\label{fig:bad-sit-diagram}
\end{figure}

>>>>

Our optimizer searches (\Cref{sec:planning}) a restricted space (\Cref{sec:search-space}) of query plans containing imputation operators (\Cref{sec:operators}) for an optimal plan given an estimate of the runtime of the query, information lost through imputation operations, and a parameter trading off the two.

This plan is subject to two constraints (\Cref{sec:placement}).
First, it must not emit any tuples which contain null values, regardless of the state of the base tables.
Second, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) must never observe a null value in any attribute that they operate on directly.
Our planning algorithm is agnostic to the type of imputation used (\Cref{sec:imputation}), as the properties of the imputation are summarized by the cost model (\Cref{sec:cost-model}).

\subsection{Imputation operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all null values with non-null values for attributes in $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
\textit{Drop} simply removes from $R$ all tuples which have a null value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no null values for
attributes in $C$.

@@

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
\Cref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.

@@

The computation of $\textsc{Time}(q)$ and $\textsc{Loss}(q)$ relies on having an accurate cardinality estimate of a query plan $q$ and its sub-plans. 
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
\Cref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

The time computation for \textit{Impute} depends on the properties of the underlying
algorithm. \ProjectName{} treats the actual machine learning algorithm as a black box
and simply queries it for a time estimate. In the base implementation of our system,
we use an iterative algorithm (see~\Cref{sec:imputation}), where the runtime cost
is a function of the number of attributes with no missing data, number of attributes with
missing data being imputed, number of tuples, and the number of iterations for the algorithm.

@@

The query planner must select a join ordering in addition to placing imputation operators as described in~\Cref{sec:placement}.

Before describing the operation of the query planner, we describe the semantics of a specialized data structure for holding sub-plans, called a \emph{plan cache}.
At a high level, a plan cache is a mapping from a set of tables to a set of dominating plans that perform the required joins and filters over these tables.
We store multiple plans rather than a single plan because each plan has a distinct set of dirty attributes and a choice over the plan frontier
can only be made once the set is final. Details are shown in~\Cref{fig:semantics-plan-cache}.

@@

\Cref{algo:plan-helpers} presents a series of helper functions used by the top-level planner, shown in~\Cref{algo:top-level-planner}.

The planner operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}-\Cref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

@@

\Cref{theorem:no-bad-sit} states that there is no \textit{bad situation} such as that depicted in~\Cref{fig:bad-sit-diagram}

@@

% arrows
\draw[->] (p1) -- node[right] {$\scriptstyle I$} (p1');
\draw[->] (p2) -- node[right] {$\scriptstyle I$} (p2');
\end{tikzpicture}
\caption{A bad Pareto frontier selection would choose $p_2$ over $p_1$, as it is locally optimal, but with the available transformations ($I$) $p_2$  can only reach $p_2'$, which is dominated by the inaccessible $p_1'$.~\Cref{theorem:no-bad-sit} says such a situation cannot arise when constructing Pareto frontiers on a dirty-set basis.}
\label{fig:bad-sit-diagram}
\end{figure}

### 2017-02-26 paper/experiments.tex
paper: \ref to \Cref for standardized section names

The original tables have a large number of attributes, in some cases providing more granular
tests results or alternative metrics. In order to simplify the presentation and
exploration of queries, we focused on a subset of the attributes for each table.
Table~\ref{table:nhanes-description} shows the attributes selected, along with the
percentage of nulls. It is worth noting that for readability, we have replaced the
NHANES variable names with self-explanatory attribute names.

@@

For our purposes, we use a version of the data that has been 
pre-processed, but where plenty of missing values remain. It is
worth noting that given the table design (where many attributes
are related), many of these missing values are to be
expected. The dataset we started with had 15,620 rows and 113
columns. We chose 17 attributes, which are outlined in Table~\ref{table:fcc-description}, along with the percentage of missing values.

@@

\subsection{Queries}
We collect a set of queries (Table~\ref{fig:queries}) that we think are both representative,
in that they could reasonably be written by a user in the course of data analysis, and
interesting to plan.

@@

Queries 1-4 are on the CDC data. In Query~\ref{q1}, we calculate the average height for
users based on their income data. In Query~\ref{q2}, we compare cholesterol levels for
individuals with low, medium, and high incomes spectrum and above a certain weight. In
Query~\ref{q3}, we extract the maximum blood lead levels for children under the age of 6
years. In Query~\ref{q4}, we calculate the average systolic blood pressure, by gender, for
subjects with a body mass index indicating obesity.

Queries 5-8 are on the FCC data. In Query~\ref{q6}, we calculate the average income for
survey participants, based on their bootcamp attendance. In Query~\ref{q7} estimates the
average age of women from the United States who participated. In Query~\ref{q8}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp.

@@

Figure~\ref{fig:runtimes} provides a summary of the performance results. The first plot
shows average running times for \ProjectName{} plans in two configurations:
quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
running times required for each query when performing imputation at the base tables.  In all
cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
with performance improvement on a factor of ten. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
\caption{Query runtimes with \ProjectName{}}
\label{subfig:project-runtime-queries}
\end{subfigure}
~
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
\caption{Query runtimes with imputation on base tables}
\label{subfig:base-runtime-queries}
\end{subfigure}
\caption{Comparing \ProjectName{} runtimes for each query
    (\ref{subfig:project-runtime-queries}) with the runtime for a query when base-table
    imputations are used (\ref{subfig:base-runtime-queries})}
\label{fig:runtimes}
\end{figure}

Figure~\ref{fig:plantimes} provides a summary of the planning times for each of the queries.
We exclude the planning time for queries that impute at base table, as that requires no
planning. In the median query across a variety of queries and choices of $\alpha$, planning
constituted 8.5 percent of total runtime by \ProjectName{}. The one-standard-deviation
intervals around the mean planning time often overlap, suggesting the planning component is
constant in $\alpha$.

@@

Table~\ref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on the cleaned data. Each query was run 200 iterations in both settings
and then results from the two approaches were paired up and compared tuple-wise. We average
tuple-wise absolute percentage deviations within each iteration of a query, and we report
this value averaged over all iterations.  We can see that optimizing for quality indeed
reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
approach are low in all cases, indicating that on-the-fly imputation produces similar
results to imputation at the base tables.

@@

In many real-world cases, applying the imputation step at the base table is prohibitive.
To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}).
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.

>>>>

The original tables have a large number of attributes, in some cases providing more granular
tests results or alternative metrics. In order to simplify the presentation and
exploration of queries, we focused on a subset of the attributes for each table.~\Cref{table:nhanes-description}
shows the attributes selected, along with the
percentage of nulls. It is worth noting that for readability, we have replaced the
NHANES variable names with self-explanatory attribute names.

@@

For our purposes, we use a version of the data that has been 
pre-processed, but where plenty of missing values remain. It is
worth noting that given the table design (where many attributes
are related), many of these missing values are to be
expected. The dataset we started with had 15,620 rows and 113
columns. We chose 17 attributes, which are outlined in~\Cref{table:fcc-description}, along with the percentage of missing values.

@@

\subsection{Queries}
We collect a set of queries (\Cref{fig:queries}) that we think are both representative,
in that they could reasonably be written by a user in the course of data analysis, and
interesting to plan.

@@

Queries 1-4 are on the CDC data. In~\Cref{q1}, we calculate the average height for
users based on their income data. In~\Cref{q2}, we compare cholesterol levels for
individuals with low, medium, and high incomes spectrum and above a certain weight. In~\Cref{q3}, 
we extract the maximum blood lead levels for children under the age of 6
years. In~\Cref{q4}, we calculate the average systolic blood pressure, by gender, for
subjects with a body mass index indicating obesity.

Queries 5-8 are on the FCC data. In~\Cref{q5}, we calculate the average income for
survey participants, based on their bootcamp attendance. In~\Cref{q6} estimates the
average age of women from the United States who participated. In~\Cref{q7}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In~\Cref{q8}, we join the FCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp.

@@

\Cref{fig:runtimes} provides a summary of the performance results. The first plot
shows average running times for \ProjectName{} plans in two configurations:
quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
running times required for each query when performing imputation at the base tables.  In all
cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
with performance improvement on a factor of ten. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_imputedb.png}
\caption{Query runtimes with \ProjectName{}}
\label{subfig:project-runtime-queries}
\end{subfigure}
~
\begin{subfigure}{\linewidth}
\includegraphics[width=\columnwidth]{figures/running_times_base_tables.png}
\caption{Query runtimes with imputation on base tables}
\label{subfig:base-runtime-queries}
\end{subfigure}
\caption{Comparing \ProjectName{} runtimes for each query
    (\Cref{subfig:project-runtime-queries}) with the runtime for a query when base-table
    imputations are used (\Cref{subfig:base-runtime-queries})}
\label{fig:runtimes}
\end{figure}

\Cref{fig:plantimes} provides a summary of the planning times for each of the queries.
We exclude the planning time for queries that impute at base table, as that requires no
planning. In the median query across a variety of queries and choices of $\alpha$, planning
constituted 8.5 percent of total runtime by \ProjectName{}. The one-standard-deviation
intervals around the mean planning time often overlap, suggesting the planning component is
constant in $\alpha$.

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for
\ProjectName{}'s query results when compared to running imputation on the base tables and
executing the query on the cleaned data. Each query was run 200 iterations in both settings
and then results from the two approaches were paired up and compared tuple-wise. We average
tuple-wise absolute percentage deviations within each iteration of a query, and we report
this value averaged over all iterations.  We can see that optimizing for quality indeed
reduces the SMAPE of query results. In general, the SMAPE relative to the base-imputation
approach are low in all cases, indicating that on-the-fly imputation produces similar
results to imputation at the base tables.

@@

In many real-world cases, applying the imputation step at the base table is prohibitive.
To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
aggregate query over the ACS dataset in our final experiment(\Cref{query-acs}).
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.

### 2017-02-26 paper/introduction.tex
paper: \ref to \Cref for standardized section names

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
queries that execute in a fraction of the time it takes to impute base tables and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the base-table imputation approach, as indicated by low symmetric-Mean-Absolute-Percentage-Errors in our empirical results
over real-world data sets (see Sec.~\ref{sec:experiments} for details).

>>>>

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
queries that execute in a fraction of the time it takes to impute base tables and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the base-table imputation approach, as indicated by low symmetric-Mean-Absolute-Percentage-Errors in our empirical results
over real-world data sets (see~\Cref{sec:experiments} for details).
